---
title: "Reproducible Research: Peer Assessment 1"
author: "Andreas Schmidt Jensen"
output: 
  html_document:
    keep_md: true
---

# Explorations in dataset on activity measured in 5-minute intervals.

First, initialize by loading required libraries.
```{r init}
library(chron)
library(dplyr)
library(ggplot2)

# set timezone to UTC to make Chron plot time-series correctly.
Sys.setenv(TZ='UTC')
```

## Loading and preprocessing the data
We read the data-file and convert the `date` column to the type Date, which makes time-based analysis easier.
```{r loading data}
df <- read.csv("activity.csv", sep=",")
df$date <- as.Date(df$date, "%Y-%m-%d")
df$interval <- times(sub(x=sprintf("%04d", df$interval), pattern='([0-9]{2})([0-9]{2})', replacement='\\1:\\2:00'))
```
We see that the dataframe consists of `r nrow(df)` rows with `r ncol(df)` variables and provide a brief summary of the data:
```{r summary}
str(df)
summary(df)
```

## What is mean total number of steps taken per day?
First, we create a histogram of the total number of steps taken per day. We ignore any missing values.
```{r histogram_-_per_day}
per_day <- na.omit(df) %>% group_by(date) %>% summarise(steps = sum(steps))
hist(per_day$steps, breaks=20)
abline(v = median(per_day$steps), lwd=3, col="red")
```

We see that the median seems to be at around 10000 steps (actually plotted as the red line).

We calculate the mean and median total number of steps taken per day:
```{r mean and median per day}
mean(per_day$steps)
median(per_day$steps)
```

## What is the average daily activity pattern?
We group the data by each 5-minute interval and calculate the average number of steps for each interval. We ignore any missing values.
This data can then be plotted to show the average daily activity pattern.
```{r average_within_day}
per_interval <- na.omit(df) %>% group_by(interval) %>% summarise(steps = mean(steps))
ggplot(per_interval, aes(x = interval, y = steps), xlab="Interval", ylab="Average number of steps") + geom_line() + scale_x_chron(format="%H:%M", n = 10)
```

We see a peak just before 9:00, but we can extract the actual interval using `which.max`:
```{r max interval}
per_interval[which.max(per_interval$steps),]
```

## Imputing missing values

We see that the dataset has a number of missing values:
```{r number of missing values}
sum(is.na(df$steps))
```
This corresponds to quite a large percentage of total values that are missing:
```{r percentage of missing values}
mean(is.na(df$steps))*100
```

We impute the missing values using the mean for that specific 5-minute interval.
```{r imputing with mean of 5-minute interval}
df.imputed <- df
# simple for-loop going through each row, imputing if the steps-value is missing.
for (i in 1:nrow(df.imputed)) {
  df.imputed[i,"steps"] <- ifelse(is.na(df.imputed[i,"steps"]), per_interval[per_interval$interval==df.imputed[i,"interval"],"steps"], df.imputed[i,"steps"])
}

```

We can now verify that there are no missing values in the dataset.
```{r verification of imputation}
sum(is.na(df.imputed))
```

### Histogram - imputed values
We can create a histogram for the imputed values, to see what the imputation does to the distribution.
```{r histogram_-_imputed}
per_day <- df.imputed %>% group_by(date) %>% summarise(steps = sum(steps))
hist(per_day$steps,  breaks = 20)
abline(v = median(per_day$steps), lwd=3, col="red")
```

We see that the median seem to be around the same value as before, but we can calcuate both the median and the mean:
```{r mean and median - imputed}
mean(per_day$steps)
median(per_day$steps)
```

We see that mean is actually unchanged but the median has moved a bit. Since we're imputing using the mean, it is not a surprising result.

## Are there differences in activity patterns between weekdays and weekends?
Now let's investigate if there are any difference in activity patterns during the weekdays and in the weekend. We first create a factor-column in the imputed dataset containing either 'weekday' or 'weekend', depending on the date. 
```{r add weekday factor}
wds <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
df.imputed$weekday <- factor((weekdays(df.imputed$date) %in% wds), levels=c(F, T), labels=c('weekend', 'weekday'))
```

We can then create a panel plot to see if there are any differences in the activity patterns.

```{r plot_panel_for_weekday_and_weekend}
per_interval <- df.imputed %>% group_by_at(vars(interval, weekday)) %>% summarise(steps = mean(steps))
ggplot(per_interval, aes(x = interval, y = steps), xlab="Interval", ylab="Average number of steps") + 
  geom_line() + 
  scale_x_chron(format="%H:%M") +
  facet_grid(weekday ~ .)
```

We see that the peak for weekends is lower, but the mean seems to be higher, which we can verify:


```{r verify different in mean}
mean(per_interval[per_interval$weekday=='weekend',]$steps)
mean(per_interval[per_interval$weekday=='weekday',]$steps)
```
The average number of steps is indeed higher in weekends compared to weekdays.

